The Controlled Choice Design and Paternalism in Pawnshop Borrowing

### 2023-08-29 OA.tex
Implementation of validity tot_tut

\todo[inline]{Add exercise about falsification of exclusion that Frank suggested.}
The \cite{huber_mellace} bounds for the TOT \& TUT APR are respectively:
\[\text{TOT} : [-366, 490]\quad\quad\quad\text{TUT} : [-19, 109]\]

Since both TUT (24) and ToT (121) estimates from the LATE approach lies inside these bounds, we cannot falsify the exclusion restriction.

>>>>

The \cite{huber_mellace} bounds for the TOT \& TUT APR are respectively:  $[-366, 490]$, and $[-19, 109]$.
Since both TUT (24) and ToT (122) estimates from the LATE approach lies inside these bounds, we cannot falsify the exclusion restriction. Moreover, the p-values associated with testing violation of inequalities \ref{eq:testable0}, and \ref{eq:testable1} is equal to one in both cases\footnote{We include the implementation of the validity test for the exclusion restriction for choosers and non-choosers in our STATA package.}.

### 2023-08-29 Tables/transcribed.tex
table ss with pre-experiment data
Implementation of validity tot_tut

      &  \\
      &  \\
3     & \textbf{How much would you sell the item you want to pawn for?       \_\_\_\_\_\_\_\_\_\_\_\_\_\_ pesos} \\

@@

17    & \textbf{If you are saving money and a family member wants to use it for something } \\
      & (a) I would only give him the money for an urgent expenze \\
      & (b) I would give him the money even if it was not an urgent expense \\
      & (c) I would not give him/her the money regardless \\
      & (d) No one would ask me for my money \\

@@

19    & \textbf{Do you have other items you could pawn?} \\

>>>>

3     & \textbf{How much do you think the item you plan to pawn is worth?       \_\_\_\_\_\_\_\_\_\_\_\_\_\_ pesos} \\

@@

17    & \multicolumn{1}{p{54.91em}}{\textbf{If you are saving money and a family member wants to use it for something }} \\
      & \multicolumn{1}{p{54.91em}}{(a) I would only give him the money for an urgent expenze} \\
      & \multicolumn{1}{p{54.91em}}{(b) I would give him the money even if it was not an urgent expense} \\
      & \multicolumn{1}{p{54.91em}}{(c) I would not give him/her the money regardless} \\
      & \multicolumn{1}{p{54.91em}}{(d) No one would ask me for my money} \\

@@

19    & \multicolumn{1}{p{54.91em}}{\textbf{Do you have other items you could pawn?}} \\

### 2023-08-29 donde.tex
Implementation of validity tot_tut

Making a case for private paternalism is not easy. It involves comparing the causal effect of forcing a certain product characteristic against the effect of letting people choose it. Because choice may be correlated with unobservables, identifying these two effects is challenging in observational studies.  %This is not a straightforward task because choice will be driven by unobservable factors, and to address the question fully 
Moreover, painting a complete picture of the benefits of forcing requires counterfactuals for both choosers and non-choosers alike, since the differential benefit of forcing arises from impacts among those who do not freely choose commitment. While a large behavioral literature has illustrated that allowing borrowers or savers to choose commitment-style products increase savings in an intention-to-treat (ITT) sense \citep{thaler2004save, prina2015banking, brune2016facilitating, callen2019headwaters, Pascaline, Ashraf}, this literature has rarely addressed the benefits of imposing commitment or its differential impacts on those who would not independently choose commitment.  
%\footnote{Selection on gains has such wide acceptance that economists often impose it in their models} (Although the examples above are plausible, substantiating a case for private paternalism is not easy. It involves comparing the causal effect of forcing a certain product characteristic versus the effect of letting people choose on that characteristic. This is not an easy task because a given person wither has or does not have choice, and because choice induces selection, potentially biasing causal estimates. A large behavioral literature has illustrated that allocating borrowers to a commitment-savings product increase savings \citep{thaler2004save, prina2015banking, brune2016facilitating, callen2019headwaters, Pascaline, Ashraf}, but this literature does not show that commitment would not be chosen by those that would benefit from it, that is, that the treatment effect on the untreated (TUT) is positive. 
Indeed, selection on gains has such wide acceptance that economists often impose it as a part of Roy models on topics as diverse as migration \citep{borjas1989economic}, job search \citep{lippman1976economics}, and program evaluation \citep{bjorklund1987estimation}. In microfinance, a context close to ours, \citep{Murdoch} claim that there exists positive selection, citing a positive correlation between displaying present biased preferences in surveys and selecting into microfinance as evidence of borrowers' actually demanding the discipline imposed by frequent payments. However, they are unable to show that borrowers indeed benefit or that they indeed positively select based on treatment effects.

@@

Two other recent papers also shed some light on private paternalism. In the context of food choice, \cite{Sprenger} elicit dynamic inconsistencies and show that demand for commitment is negatively correlated with these inconsistencies. In the context of school choice, \cite{Walters} finds that students who select into more effective schools have smaller treatment effects from attending than those who do not.
Our paper differs from these in both its empirical setting and its methodology.
As far as we are aware, ours is the first experiment to combine control, forced treatment, and treatment choice arms and explain how such a design can be used to point identify TUT effects and a range of other interesting causal parameters under minimal assumptions.
We call this approach the ``Controlled Choice design'' for short.
Other approaches to going beyond local average treatment effects (LATE) from the literature either assume no unobserved selection-on-gains \citep{aronow2013beyond,angrist2013extrapolate} (the ``LATE-and-reweight'' approach) or rely on a combination of instruments with rich support and additional structural assumptions such as additive separability of observed and unobserved determinants of treatment effect heterogeneity \citep{heckman2007econometric, cornelissen2018benefits} (the marginal treatment effects approach).
\cite{Walters} takes the latter approach, combining a distance instrument with additional structural assumptions, obtaining model-based TUT and ATE estimates that differ substantially from those implied by a LATE-and-reweight approach. 
In contrast, we leverage our experimental design to point identify the TUT effect using a binary instrument without the need for a structural model, relying instead on relatively simple exclusion restrictions. This restriction has testable implications that we fail to reject. Rather than assuming away unobserved selection-on-gains, as in the LATE-and-reweight approach, we estimate it directly.
Unlike \cite{Sprenger} we directly identify the TUT, obviating the need to first elicit preferences before testing for negative selection.

@@

%Above we used outcome data from the forced arms ($Z_i\neq 2$) along with the commitment take-up rate in the choice arm ($Z_i = 2$) to argue that at least some of the \hl{89\%} of participants who did \emph{not} choose commitment would have been better off if they had. 
%By additionally examining \emph{outcome data} in the choice arm, it is possible to say more. 
We now show how our experimental design, henceforth the ``controlled choice design,'' combined with the exclusion restrictions from \eqref{eq:exclusion0} and \eqref{eq:exclusion1} can be used to point identify a number of interesting and economically-relevant causal quantities without the need for additional structural restrictions.
First we identify the treatment on the treated (TOT) and treatment on the untreated (TUT) effects, defined as follows:
\[
\text{TOT} \equiv \mathbbm{E}(Y_{i1} - Y_{i0} | C_i = 1), \quad
\text{TUT} \equiv \mathbbm{E}(Y_{i1} - Y_{i0} | C_i = 0).
\]
%The TuT and TOT effects are particularly interesting for understanding the importance of paternalism. 
If the TUT is positive, this means that those who did not choose commitment would have experienced better outcomes, \emph{on average}, if they had. 
In a canonical Roy model, the TOT should exceed both the TUT and average treatment effect (ATE).
If the TOT is statistically distinguishable from and substantially larger than the TUT, this provides empirical support for the relevance of selection-on-gains in real-world decision-making.
Because our design identifies all three quantities (as shown in Appendix \ref{append:randchoice}), it allows us to test this implication directly and to calculate the average selection on gains (ASG):
\[
\text{ASG} \equiv \mathbbm{E}[Y_{i1} - Y_{i0}|C_i = 1] - \mathbb{E}[Y_{i1} - Y_{i0} | C_i = 0] = \text{TOT} - \text{TUT}.
\]
The controlled choice design also identifies both the average selection bias (ASB) and the average selection on levels (ASL). 
\[
\text{ASB} \equiv \mathbbm{E}[Y_{i0}|C_i = 1] - \mathbbm{E}[Y_{i0}|C_i = 0], \quad 
\text{ASL} \equiv \mathbbm{E}[Y_{i1}|C_i = 1] - \mathbbm{E}[Y_{i1}|C_i = 0].
\]
A companion STATA package accompanying this paper provides estimators of the TOT, TUT, ASG, ASB, and ASL, along with cluster-robust standard errors for each.\footnote{See Appendix \ref{append:randchoice} for more details.}

@@

Under the assumption that our instrumental forest estimates capture the main sources of treatment effect heterogeneity, we can use them to assess whether particular borrowers in the choice arm made ``mistakes'' in their decision to accept or refuse the commitment contract, in terms of predicted financial costs of the loan.
To do this we use the same information that is depicted in Figure \ref{heterogeneous_effects}, but present it in a different way.
For each non-chooser in the choice arm, we use our instrumental forest from above to estimate the conditional TUT effect, given her observed covariates.
Of course conditional average effects need not equal individual treatment effects, and our APR outcome may not capture all of the costs and benefits that are relevant for individual borrowers' decisions.
To account for this, we define a ``mistake'' for a non-chooser to be a conditional TUT estimate that significantly exceeds some large and positive APR threshold, e.g.\ 25\%. 
At any such threshold, we can calculate the percentage of non-choosers in the choice arm who made a ``mistake'' by not choosing commitment.   
(If treatment it beneficial, it should be chosen.)
The results of this exercise can be read off from the green curve in Figure \ref{choose_wrong}.
Defining $F_{\text{TUT}}(\delta)$ to be the CDF corresponding to the density of conditional TUT estimates from Figure \ref{heterogeneous_effects}, the green curve in Figure \ref{chose_wrong} is merely $[1 - F_{\text{TUT}}(\delta)] \times 100\%$.
In other words, the green curve gives the percentage of non-choosers who made a ``mistake'' when mistakes are defined at a particular APR threshold.
The green shaded region gives associated 95\% pointwise confidence bounds.

>>>>

Making a case for private paternalism is not easy. It involves comparing the causal effect of forcing a certain product characteristic against the effect of letting people choose it. Because choice may be correlated with unobservables, identifying these two effects is challenging in observational studies.  %This is not a straightforward task because choice will be driven by unobservable factors, and to address the question fully 
Moreover, painting a complete picture of the benefits of forcing requires counterfactuals for both choosers and non-choosers alike, since the differential benefit of forcing arises from impacts among those who do not freely choose commitment. While a large behavioral literature has illustrated that allowing borrowers or savers to choose commitment-style products increase savings in an intention-to-treat (ITT) sense \citep{thaler2004save, prina2015banking, brune2016facilitating, callen2019headwaters, Pascaline, Ashraf}, this literature has rarely addressed the benefits of imposing commitment or its differential impacts on those who would not independently choose commitment.  
%\footnote{Selection on gains has such wide acceptance that economists often impose it in their models} (Although the examples above are plausible, substantiating a case for private paternalism is not easy. It involves comparing the causal effect of forcing a certain product characteristic versus the effect of letting people choose on that characteristic. This is not an easy task because a given person wither has or does not have choice, and because choice induces selection, potentially biasing causal estimates. A large behavioral literature has illustrated that allocating borrowers to a commitment-savings product increase savings \citep{thaler2004save, prina2015banking, brune2016facilitating, callen2019headwaters, Pascaline, Ashraf}, but this literature does not show that commitment would not be chosen by those that would benefit from it, that is, that the treatment effect on the untreated (TUT) is positive. 
Indeed, selection on gains has such wide acceptance that economists often impose it as a part of Roy models on topics as diverse as migration \citep{borjas1989economic} and job search \citep{lippman1976economics}.
%, and program evaluation \citep{bjorklund1987estimation}. 
In microfinance, a context close to ours, \citep{Murdoch} claim that there exists positive selection, citing a positive correlation between displaying present biased preferences in surveys and selecting into microfinance as evidence of borrowers' actually demanding the discipline imposed by frequent payments. However, they are unable to show that borrowers indeed benefit or that they indeed positively select based on treatment effects.

@@

Two other recent papers also shed some light on private paternalism. In the context of food choice, \cite{Sprenger} show that individuals with the most time-inconsistent preferences are actually least likely to demand commitment, suggesting a lack of awareness as to self-control problems. In the context of school choice, \cite{Walters} finds that students who select into more effective schools have smaller treatment effects from attending than those who do not.
Our paper differs from these in both its empirical setting and its methodology.
As far as we are aware, ours is the first experiment to combine control, forced treatment, and treatment choice arms and explain how such a design can be used to point identify TUT effects and a range of other interesting causal parameters under minimal assumptions.
While \cite{fowlie2021default} likewise exploited a three-armed experimental design in their study of the effect of electricity pricing, they identified two TOT effects for different groups of ``treated'' households, whereas we simultaneously identify the TOT and TUT effects defined with respect to a single ``treated'' group of borrowers.
We call our approach the ``Controlled Choice design'' for short.
Other approaches to going beyond local average treatment effects (LATE) from the literature either assume no unobserved selection-on-gains \citep{aronow2013beyond,angrist2013extrapolate} (the ``LATE-and-reweight'' approach) or rely on a combination of instruments with rich support and additional structural assumptions such as additive separability of observed and unobserved determinants of treatment effect heterogeneity \citep{heckman2007econometric, cornelissen2018benefits} (the marginal treatment effects approach).
\cite{Walters} takes the latter approach, combining a distance instrument with additional structural assumptions, obtaining model-based TUT and ATE estimates that differ substantially from those implied by a LATE-and-reweight approach. 
In contrast, we leverage our experimental design to point identify the TUT effect using a binary instrument without the need for a structural model, relying instead on relatively simple exclusion restrictions. This restriction has testable implications that we fail to reject. Rather than assuming away unobserved selection-on-gains, as in the LATE-and-reweight approach, we estimate it directly.
Unlike \cite{Sprenger} we directly identify the TUT, obviating the need to first elicit preferences before testing for negative selection.
In a returns-to-education setting, \cite{oreopoulos2006estimating}, uses a similar direct approach to identify a LATE \emph{nearly} equals a TUT effect: the 1944 Butler Act increased the share of British children who remained in school until age 15 from 43\% to over 90\%. Because 100\% of borrowers in our forced arm are treated, we identify the TUT rather than an approximation to it.

@@

%Above we used outcome data from the forced arms ($Z_i\neq 2$) along with the commitment take-up rate in the choice arm ($Z_i = 2$) to argue that at least some of the \hl{89\%} of participants who did \emph{not} choose commitment would have been better off if they had. 
%By additionally examining \emph{outcome data} in the choice arm, it is possible to say more. 
We now show how our experimental design, henceforth the ``controlled choice design,'' combined with the exclusion restrictions from \eqref{eq:exclusion0} and \eqref{eq:exclusion1} can be used to point identify a number of interesting and economically-relevant causal quantities without the need for additional structural restrictions.
First we identify the treatment on the treated (TOT) and treatment on the untreated (TUT) effects, defined as follows:
\[
\text{TOT} \equiv \mathbbm{E}(Y_{i1} - Y_{i0} | C_i = 1), \quad
\text{TUT} \equiv \mathbbm{E}(Y_{i1} - Y_{i0} | C_i = 0).
\]
%The TuT and TOT effects are particularly interesting for understanding the importance of paternalism. 
If the TUT is positive, this means that those who did not choose commitment would have experienced better outcomes, \emph{on average}, if they had. 
In a canonical Roy model, the TOT should exceed both the TUT and average treatment effect (ATE).
If the TOT is statistically distinguishable from and substantially larger than the TUT, this provides empirical support for the relevance of selection-on-gains in real-world decision-making.
Because our design identifies all three quantities (as shown in Appendix \ref{append:randchoice}), it allows us to test this implication directly and to calculate the average selection on gains (ASG):
\[
\text{ASG} \equiv \mathbbm{E}[Y_{i1} - Y_{i0}|C_i = 1] - \mathbb{E}[Y_{i1} - Y_{i0} | C_i = 0] = \text{TOT} - \text{TUT}.
\]
The controlled choice design also identifies both the average selection bias (ASB) and the average selection on levels (ASL). 
\[
\text{ASB} \equiv \mathbbm{E}[Y_{i0}|C_i = 1] - \mathbbm{E}[Y_{i0}|C_i = 0], \quad 
\text{ASL} \equiv \mathbbm{E}[Y_{i1}|C_i = 1] - \mathbbm{E}[Y_{i1}|C_i = 0].
\]
A companion STATA package accompanying this paper provides estimators of the TOT, TUT, ASG, ASB, and ASL, along with cluster-robust standard errors for each, and a test for the validity of the exclusion restriction following \cite{huber_mellace} \footnote{See Appendix \ref{append:randchoice} for more details.}.

@@

Under the assumption that our instrumental forest estimates capture the main sources of treatment effect heterogeneity, we can use them to assess whether particular borrowers in the choice arm made ``mistakes'' in their decision to accept or refuse the commitment contract, in terms of predicted financial costs of the loan.
To do this we use the same information that is depicted in Figure \ref{heterogeneous_effects}, but present it in a different way.
For each non-chooser in the choice arm, we use our instrumental forest from above to estimate the conditional TUT effect, given her observed covariates.
Of course conditional average effects need not equal individual treatment effects, and our APR outcome may not capture all of the costs and benefits that are relevant for individual borrowers' decisions.
To account for this, we define a ``mistake'' for a non-chooser to be a conditional TUT estimate that significantly exceeds some large and positive APR threshold, e.g.\ 25\%. 
At any such threshold, we can calculate the percentage of non-choosers in the choice arm who made a ``mistake'' by not choosing commitment.   
(If treatment it beneficial, it should be chosen.)
The results of this exercise can be read off from the green curve in Figure \ref{choose_wrong}.
Defining $F_{\text{TUT}}(\delta)$ to be the CDF corresponding to the density of conditional TUT estimates from Figure \ref{heterogeneous_effects}, the green curve in Figure \ref{choose_wrong} is merely $[1 - F_{\text{TUT}}(\delta)] \times 100\%$.
In other words, the green curve gives the percentage of non-choosers who made a ``mistake'' when mistakes are defined at a particular APR threshold.
The green shaded region gives associated 95\% pointwise confidence bounds.
